Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-22 analysis/paper/paper.Rmd
added references and finished tthe review of GMM studies

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons[@Birch2019; @Buchanan2019; @Cardillo2010; @Lycett2013; @Selden2018], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017]. The applications of GMM couple with multivariate statistics for shape variables of artifacts are commonly associated with topics of topological evolution or variations in assemblages to identify groups [@Doyon2019; @Eren2015; @Selden2018].

In addition to vibrations, shape standardization of artifacts and its causes is another topic. A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape. This enables the identification of clustered analytical groups through visualization for the results, followed by statistical tests on the shape variables. In this way, we can explore whether there is less variations or high standardization in the same assemblage over time or between assemblages over larger geographical areas. For example, @Archer2015's case study of Still Bay points in Southern Africa suggests an increase in point shape standardization over time that may relate to a maintenance of finished points. For comparison between several sites, @Buchanan2018 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting. Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel. There are also other factors could lead to an increase in standardization of point shapes such as low level of cultural innovation in a small group [@Okumura2014].

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approach, Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984], and has been increasing applied to different archaeological materials. Most applications are associated with human remains and animal bones to examine biological variation, and also stone artifacts to explore stone tool topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].

Despite the EFA applications for ceramics is rare, this approach is promising for answering questions related to ceramic taxonomy [@Topi2017; @Wilczek2014]. For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region. EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation. In such cases, the focus is the whole shape instead of the different components of a shape. We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization. The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.

>>>>

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017]. With univariate or multivariate statistics for shape data generated by GMM, the application to artifacts can answer research questions related to topological evolution or variations in assemblages to define groups [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova]. A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape. This enables the identification of clustered analytical groups through the visualization of results, followed by statistical tests to robustly distinguish them.

One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas. There is many discussion about lithic technology using multivariate analysis on shape variables or coefficient of variations on associated metric data. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points. With considerations of a set of metric data, @Buchanan2018; @Buchanan2019 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting. Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel. Other factor, such as low level of cultural innovation in a small group could also lead to an increase in standardization of point shapes [@Okumura2014]. For ceramics, @Topi2017 found a standardized vessel shape from the Casas Grandes culture in northwest Mexico using coefficient of variations for the positions of semi-landmarks across shape groups. They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity.

Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001]. In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity. Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020]. More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007]

Despite the wide use of landmark approach in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that demonstrated the changes in forms from a more angular to rounded shape over time. Since then, EFA has been applied to different archaeological materials. 
Similar to landmark approach, most applications are associated with human remains and animal bones to examine biological variation, and stone artifacts to explore topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].

The EFA applications for ceramics is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization. For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region. EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation. In such cases, the focus is the whole shape instead of the different components of a shape. We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization. The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
